---
title: "LCFS Analysis"
author: "U2567455"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  word_document:
    toc: true
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    code_folding: show
  pdf_document:
    toc: true
editor_options:
  markdown:
    wrap: 72
---

# Initial Setup

Please keep this file in the `Assessment` directory alongside
`data_dict_mod.csv` and the `UKDA-9603-stata` directory. Individual R
scripts can be also used from repo github/.

The full data set could be downloaded from:
<https://datacatalogue.ukdataservice.ac.uk/studies/study/9603>

-Please unzip the file and, (optional) the data dictionary folders
inside

-Please Remember to set the working directory

Alternatively, clone from <https://github.com/u2567455/living-cost-food-survey-2026>

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = TRUE,
  fig.width = 8,
  fig.height = 5,
  fig.align = "center",
  dpi = 120
)

required_packages <- c(
  "haven", "labelled", "tidyverse", "rpart", "rpart.plot",
  "randomForest", "gbm", "patchwork", "caret", "knitr", "rmarkdown", "class",
  "ggplot2", "ggthemes"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages) > 0) {
  stop(
    "Install the required packages before knitting: install.packages(c(",
    paste(sprintf('"%s"', missing_packages), collapse = ", "),
    "))"
  )
}
```

## Libraries and report styles

The shared plotting style and colors are global

```{r message=FALSE}
library(haven)
library(labelled)
library(tidyverse)
library(rpart)
library(rpart.plot)
library(randomForest)
library(gbm)
library(patchwork)
library(caret)
library(pROC)
library(class)
library(ggplot2)
library(ggthemes)
```

```{r report-style}
theme_colors <- c(
  "#0F172A", "#1E3A5F", "#2E5E7E",
  "#3B82A8", "#5DADE2", "#A7D3F2"
)

theme_report <- theme_bw(base_size = 12) +
  theme(
    plot.title = element_text(size = 14, face = "bold"),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 11),
    legend.title = element_text(size = 11),
    legend.text = element_text(size = 10),
    strip.text = element_text(size = 11),
    panel.grid.minor = element_blank()
  )
```

# Data preparation

## Import and initial variable selection

The raw data are read from a Stata file. Expenditure-code and
questionnaire-filter variables are removed. Please use setwd() as needed

```{r import-data}
lcfs_2024 <- read_dta("UKDA-9603-stata/stata/stata13_se/dvhh_urban_rural_uk_anon_v2_2024.dta")

cols <- names(lcfs_2024)
cat("Var count: ",length(cols),"\n")

#Refer docs > UKDA-9603-stata\mrdoc\pdf\9603_volume_g_2024_25.pdf for variable types

a_cols <- cols[startsWith(cols, "a")]
b_cols <- cols[startsWith(cols, "b")]
c_cols <- cols[startsWith(cols, "c")]
p_cols <- cols[startsWith(cols, "p")]
p_end_cols <- cols[endsWith(cols, "p")]

expenditure_cols <- grep("^c.*[0-9]", cols, value = TRUE)
important_cols <- setdiff(cols, expenditure_cols)

filter_cols_searchlist <- c(
  "geteldet", "eteldet", "tvserv", "tvfilt",
  "geteltv", "eteltv", "telserv", "telsmart", "combfilt"
)
pattern <- paste(filter_cols_searchlist, collapse = "|")
filter_cols <- grep(pattern, cols, value = TRUE)
important_cols <- setdiff(important_cols, filter_cols)

lcfs_2024_cleaned <- lcfs_2024[, important_cols]
cat("Var count after cleaning : ",length(names(lcfs_2024_cleaned)),"\n")
```

## Data dictionary and selected variables

The existing `data_dict_mod.csv` controls which variables enter the
analysis: rows marked `isUsable = 1` are selected. Lines 123-129 are
optional

```{r data-dictionary}
data_dict <- tibble(
  name = names(lcfs_2024_cleaned),
  description = unname(var_label(lcfs_2024_cleaned, unlist = TRUE)),
  isUsable = 0
)

write.csv(data_dict, file = "data_dict.csv", row.names = FALSE)

#Manually sift the important variables and create a data_dict_mod.csv from data_dict.csv
#data_dict_mod <- read_csv("data_dict_mod.csv")
#Selected variables into a variable (Toggle isUsable from .csv to include any var)
#selected_variables <- data_dict_mod %>% filter(isUsable ==1)

#Or use below list
study_variables <- c(
  "a049", "a124", "ctpcnt", "ctproxy", "g018", "g019",
  "p204", "p535cp", "p535tp", "p535p", "p537", "p538",
  "p538c", "p539", "p540", "p541", "p541c", "p542",
  "p542c", "p543", "p543c", "p544", "p544c", "p545",
  "p545c", "p546", "p546c", "p547", "p547c", "p548",
  "p548c", "p549", "p549c", "a054", "a056", "a060",
  "a091", "a093", "a094", "p203", "p304", "p336",
  "p348", "p153t", "gorx", "p365p", "p344p", "anon_income",
  "p493p", "a069p", "mcvkid", "oecdvkid", "eqincdo_updated",
  "eqincgo_updated", "urindew", "urindsc", "a1711p", "a1646p",
  "a172ap", "a172bp", "a172p", "a172cp", "a112"
)

#Selected variables
selected_variables <- data_dict %>%
  filter(name %in% study_variables)

#Add a clean variable name (remove special chars, space etc)
selected_variables <- selected_variables %>%
  dplyr::mutate(
    clean_description = description %>%
      str_trim() %>%
      str_replace_all("[^A-Za-z0-9 ]", "") %>%
      str_replace_all("\\s+", "_") %>%
      str_to_lower() %>%
      paste(name, sep = "_")
  )

lcfs_data <- lcfs_2024_cleaned[, selected_variables$name]
names(lcfs_data) <- selected_variables$clean_description
cat("Var count after selecting : ",length(names(lcfs_data)),"\n")
```

## Rename variables and handle missing values

Selected labelled variables are converted to factors, and rows
containing missing values are removed, following the preparation script.

```{r clean-data}
#Variable renames. Add everything here as needed (for clarity)
lcfs_data <- lcfs_data %>%
  dplyr::rename(
    urban_rural_indicator = urindew_urindew,
    council_tax = anonymised_council_tax_ctproxy,
    socio_economic_group = socioeconomic_group_household_reference_person_a091,
    government_office_region = government_office_region_modified_gorx,
    economic_position_of_hrp = economic_position_of_household_reference_person_a093,
    gross_normal_income_of_hrp = gross_normal_income_of_hrp_by_range_a060,
    class_of_hrp = ns_sec_8_class_of_household_reference_person_a094,
    household_size = household_size_a049,
    total_personal_expenditure = total_personal_expenditure_children_and_adults_p153t,
    food_adults = ons_food_adults_p538
  )


#Convert haven types to R
lcfs_data <- lcfs_data %>%
  mutate(across(where(is.labelled), as_factor))

#Remove NA values
rows_before_missing <- nrow(lcfs_data)
lcfs_data <- na.omit(lcfs_data)

knitr::kable(
  data.frame(
    Measure = c("Imported rows", "Rows removed for missing values",
                "Analysis rows", "Selected variables"),
    Count = c(nrow(lcfs_2024), rows_before_missing - nrow(lcfs_data),
              nrow(lcfs_data), ncol(lcfs_data))
  ),
  caption = "Data preparation summary"
)
```

# Plot Income

```{r plot-income}
the_base <- ggplot(lcfs_data, aes(x = anonincome_anon_income))
the_base

basic_histogram <- the_base +
  geom_histogram(
    aes(fill = after_stat(x)),
    bins = 20,
    colour = "#0F172A",
    linewidth = 0.4
  ) +
  scale_fill_gradientn(
    colours = theme_colors,
    guide = "none"
  ) +
  labs(
    title = "Distribution of Income",
    x = "Income",
    y = "Count"
  ) +
  theme_report
basic_histogram
```

# Training and test data

This is optional. Caret package flags variables with near zero variance
For example: Motoring expenditure by Kids

```{r remove-zero-var}
#Remove variables with near Zero Variance
nzv_cols <- nearZeroVar(lcfs_data, names = TRUE)
lcfs_data <- lcfs_data %>% select(-all_of(nzv_cols))
knitr::kable(
  data.frame(
    Set = "Near-zero variance variables removed",
    Count = length(nzv_cols)
  ),
  caption = "Near-zero variance variables removed from LCFS data"
)
```

A reproducable random split assigns 60% of the cleaned observations to
training and the remaining 40% to testing. All models use the same rows.

```{r train-test-split}
set.seed(123)
n <- nrow(lcfs_data)
train_idx <- sample(seq_len(n), size = floor(0.6 * n))
test_idx <- setdiff(seq_len(n), train_idx)
lcfs_train <- lcfs_data[train_idx, ]
lcfs_test <- lcfs_data[test_idx, ]

knitr::kable(
  data.frame(Set = c("Training", "Test"),
             Observations = c(nrow(lcfs_train), nrow(lcfs_test))),
  caption = "Shared training/test split"
)
```

# Decision tree

## Baseline model and predictions

The baseline regression tree uses a complexity parameter of `cp = 0.1`.

```{r decision-tree-fit}
tpe_tree <- rpart(
  total_personal_expenditure ~ ., data = lcfs_train, cp = 0.1
)
rpart.plot(
  tpe_tree,
  box.palette = "#A7D3F2",
  shadow.col = "#5DADE2",
  legend.x = NA
)

tpe_tree_predictions <- predict(tpe_tree, newdata = lcfs_test)
tree_mse <- mean(
  (lcfs_test$total_personal_expenditure - tpe_tree_predictions)^2
)
cat("Decision tree test MSE:", round(tree_mse, 3), "\n")
```

```{r decision-tree-predictions}
ggplot(
  data.frame(actual = lcfs_test$total_personal_expenditure,
             predicted = tpe_tree_predictions),
  aes(x = actual, y = predicted)
) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(title = "Decision Tree: Actual vs Predicted Expenditure",
       x = "Actual Total Personal Expenditure",
       y = "Predicted Total Personal Expenditure") +
  coord_fixed() +
  expand_limits(
    x = range(lcfs_test$total_personal_expenditure, tpe_tree_predictions),
    y = range(lcfs_test$total_personal_expenditure, tpe_tree_predictions)
  ) +
  theme_report
```

## Hyperparameter tuning

Finding the best complexity parameter

```{r decision-tree-complexity}
cp_values <- seq(0, 0.01, length.out = 100)
results <- data.frame(cp = numeric(), mse_train = numeric(), mse_test = numeric())

for (cp in cp_values) {
  tree_model <- rpart(
    total_personal_expenditure ~ ., data = lcfs_train,
    cp = cp, minsplit = 50, maxdepth = 20
  )
  predictions_train <- predict(tree_model, newdata = lcfs_train)
  predictions_test <- predict(tree_model, newdata = lcfs_test)
  
  mse_train <- mean((lcfs_train$total_personal_expenditure - predictions_train)^2)
  mse_test <- mean((lcfs_test$total_personal_expenditure - predictions_test)^2)
  
  results <- rbind(results, data.frame(cp = cp, mse_train = mse_train,
                                      mse_test = mse_test))
}
```

```{r decision-tree-complexity-plot, fig.width=11}
p_train <- ggplot(results, aes(x = cp, y = mse_train)) +
  geom_line(color = "#1E3A5F") +
  labs(title = "Training MSE vs Complexity Parameter",
       x = "Complexity Parameter (cp)", y = "MSE") +
  theme_report

p_test <- ggplot(results, aes(x = cp, y = mse_test)) +
  geom_line(color = "#b50279") +
  labs(title = "Test MSE vs Complexity Parameter",
       x = "Complexity Parameter (cp)", y = "MSE") +
  theme_report

p_train + p_test
```

# Random forest

## Baseline model

The initial forest uses 50 trees and samples four predictors at each
split.

```{r random-forest-baseline}
rf.lcfs <- randomForest(
  total_personal_expenditure ~ .,
  data = lcfs_train,
  ntree = 50,
  mtry = 4,
  importance = FALSE,
  na.action = na.omit
)
rf.predictions <- predict(rf.lcfs, newdata = lcfs_test)
rf.mse <- mean((lcfs_test$total_personal_expenditure - rf.predictions)^2)
cat("Baseline random forest test MSE:", round(rf.mse, 3), "\n")
```

```{r random-forest-baseline-plot}
ggplot(
  data.frame(actual = lcfs_test$total_personal_expenditure,
             predicted = rf.predictions),
  aes(x = actual, y = predicted)
) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(title = "Random Forest: Total Personal Expenditure Predictions",
       x = "Actual Total Personal Expenditure",
       y = "Predicted Total Personal Expenditure") +
  theme_report +
  coord_fixed() +
  expand_limits(
    x = range(lcfs_test$total_personal_expenditure, rf.predictions),
    y = range(lcfs_test$total_personal_expenditure, rf.predictions)
  )
```

## Hyperparameter tuning

Five-fold cross-validation on the training data compares `mtry` values
of 3, 5, and 7, and forests with 100, 300, and 500 trees.

```{r random-forest-tuning}
tune_grid <- data.frame(mtry = c(3, 5, 7))
ntree_values <- c(100, 300, 500)
rf_results_all <- data.frame()

set.seed(123)
rf_folds <- createFolds(
  lcfs_train$total_personal_expenditure, k = 5, returnTrain = TRUE
)
rf_control <- trainControl(method = "cv", number = 5, index = rf_folds)
best_rmse <- Inf
best_rf_model <- NULL

for (nt in ntree_values) {
  rf_model <- train(
    total_personal_expenditure ~ .,
    data = lcfs_train,
    method = "rf",
    metric = "RMSE",
    tuneGrid = tune_grid,
    trControl = rf_control,
    ntree = nt,
    importance = FALSE
  )
  rf_results_all <- bind_rows(
    rf_results_all,
    rf_model$results %>% mutate(ntree = nt)
  )
  model_rmse <- min(rf_model$results$RMSE)
  if (is.finite(model_rmse) && model_rmse < best_rmse) {
    best_rmse <- model_rmse
    best_rf_model <- rf_model
  }
}

if (is.null(best_rf_model)) {
  stop("Random forest tuning did not produce a model with finite CV RMSE.")
}
knitr::kable(rf_results_all, digits = 3,
             caption = "Random forest cross-validation results")
knitr::kable(
  data.frame(ntree = best_rf_model$finalModel$ntree,
             mtry = best_rf_model$bestTune$mtry, CV_RMSE = best_rmse),
  digits = 3, caption = "Selected random forest parameters"
)
```

```{r random-forest-tuning-plot}
ggplot(rf_results_all, aes(x = ntree, y = RMSE, color = factor(mtry))) +
  geom_line() +
  geom_point() +
  labs(title = "Random Forest Parameter Tuning (LCFS Data)",
       x = "Number of Trees (ntree)",
       y = "Cross-Validation RMSE", color = "mtry") +
  theme_report
```

## Tuned model test performance

```{r random-forest-test}
rf_predictions <- predict(best_rf_model, newdata = lcfs_test)
rf_mse <- mean((lcfs_test$total_personal_expenditure - rf_predictions)^2)
cat("Tuned random forest test MSE:", round(rf_mse, 3), "\n")

ggplot(
  data.frame(actual = lcfs_test$total_personal_expenditure,
             predicted = rf_predictions),
  aes(x = actual, y = predicted)
) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept = 0, slope = 1, color = "red") +
  labs(title = "Random Forest: Tuned Predictions",
       x = "Actual Total Personal Expenditure",
       y = "Predicted Total Personal Expenditure") +
  theme_report +
  coord_fixed() +
  expand_limits(
    x = range(lcfs_test$total_personal_expenditure, rf_predictions),
    y = range(lcfs_test$total_personal_expenditure, rf_predictions)
  )
```

# Gradient boosting

## Baseline model

The initial Gaussian boosting model uses 500 trees, an interaction depth
of 3, a learning rate of 0.1, and a minimum of 10 observations per
terminal node.

```{r gradient-boosting-baseline}
gb_lcfs <- gbm(
  total_personal_expenditure ~ .,
  data = lcfs_train,
  distribution = "gaussian",
  n.trees = 500,
  interaction.depth = 3,
  shrinkage = 0.1,
  cv.folds = 0,
  n.minobsinnode = 10,
  verbose = FALSE
)
gbm_train_pred <- predict(gb_lcfs, newdata = lcfs_train, n.trees = 500)
gbm_test_pred <- predict(gb_lcfs, newdata = lcfs_test, n.trees = 500)

train_mse <- mean((lcfs_train$total_personal_expenditure - gbm_train_pred)^2)
test_mse <- mean((lcfs_test$total_personal_expenditure - gbm_test_pred)^2)

knitr::kable(
  data.frame(Set = c("Training", "Test"), MSE = c(train_mse, test_mse)),
  digits = 3, caption = "Baseline gradient boosting performance"
)
```

## Hyperparameter tuning

Five-fold cross-validation compares interaction depths of 2 through 5
and learning rates of 0.01 and 0.1, retaining 500 trees and a minimum
node size of 10.

```{r gradient-boosting-tuning}
ctrl <- trainControl(method = "cv", number = 5, returnResamp = "all")
gbm_grid <- expand.grid(
  n.trees = 500,
  interaction.depth = c(2, 3, 4, 5),
  shrinkage = c(0.01, 0.1),
  n.minobsinnode = 10
)

set.seed(123)
gb_lcfs <- suppressWarnings(
  gbm_model <- train(
  total_personal_expenditure ~ .,
  data = lcfs_train,
  method = "gbm",
  metric = "RMSE",
  trControl = ctrl,
  tuneGrid = gbm_grid,
  verbose = FALSE
  )
)
print(gbm_model)
knitr::kable(gbm_model$bestTune, caption = "Selected gradient boosting parameters")
```

```{r gradient-boosting-tuning-plot}
ggplot(
  gbm_model$results,
  aes(x = interaction.depth, y = RMSE, color = factor(shrinkage))
) +
  geom_line() +
  geom_point() +
  labs(title = "Gradient Boosting Parameter Tuning (LCFS Data)",
       x = "Maximum Tree Depth", y = "Cross-Validation RMSE",
       color = "Learning Rate") +
  theme_report
```

## Tuned model test performance

```{r gradient-boosting-test}
gbm_test_pred_optim <- predict(gbm_model, newdata = lcfs_test)
test_rmse <- sqrt(mean(
  (lcfs_test$total_personal_expenditure - gbm_test_pred_optim)^2
))
cat("Tuned gradient boosting test RMSE:", round(test_rmse, 3), "\n")
```

## GBM interpretation

This returns a data frame ranked by rel.inf (relative influence, summing
to 100%)

```{r gmb-interpretation}
invisible(summary(gb_lcfs, plotit = TRUE)) #To prevent knitted doc from bloating, use invisible

#Or caret
var_imp <- varImp(gbm_model)
plot(var_imp, top = 15)

#Nice plot

# Extract variable importance
imp <- var_imp$importance
imp$Variable <- rownames(imp)

# Keep top 20
imp <- imp[order(imp$Overall, decreasing = TRUE), ]
imp <- head(imp, 20)

ggplot(imp, aes(x = reorder(Variable, Overall),
                y = Overall)) +
  geom_col(fill = "#5DADE2") +
  scale_fill_gradientn(
    colours = theme_colors,
    guide = "none"
  ) +
  coord_flip() +
  labs(
    x = "Variable",
    y = "Importance",
    title = "GBM: Variable Importance"
  ) +
  theme_report
```

# Model comparison

The table and chart compare the baseline decision tree with the tuned
random forest and tuned gradient boosting model on the same test
observations. MSE is mean squared error; RMSE is its square root. Lower
values indicate smaller prediction errors.

```{r model-comparison-table}
all_comparison <- data.frame(
  Method = c("Single Tree", "Random Forest", "GBM"),
  MSE = c(
    mean((lcfs_test$total_personal_expenditure - tpe_tree_predictions)^2),
    mean((lcfs_test$total_personal_expenditure - rf_predictions)^2),
    mean((lcfs_test$total_personal_expenditure - gbm_test_pred_optim)^2)
  )
)
all_comparison$RMSE <- sqrt(all_comparison$MSE)
knitr::kable(all_comparison, digits = 3, caption = "Test-set model comparison")
```

Plot it

```{r model-comparison-plot}
ggplot(all_comparison, aes(x = Method, y = MSE, fill = MSE)) +
  geom_col() +
  labs(title = "Model Comparison: Total Personal Expenditure",
       x = "Model", y = "Test Mean Squared Error") +
  scale_fill_gradientn(colours = theme_colors, guide = "none") +
  theme_report
```

# Logistic Regression

Logistic Regression model to predict Lowest-Income Households. Uses the
same data frame To skip tree model chunks, Please execute up to line 223
if needed.

## Data and binary outcome

Plot Income (Using equalized income)

```{r plt-icome}
#Plot a variable
# Basic histogram
# Calculate income quintile cut-points
income_quintiles <- quantile(
  lcfs_data$eqincdoupdated_eqincdo_updated,
  probs = c(0.2, 0.4, 0.6, 0.8),
  na.rm = TRUE
)

basic_histogram <- the_base +
  geom_histogram(
    aes(fill = after_stat(x)),
    bins = 20,
    colour = "#0F172A",
    linewidth = 0.4
  ) +
  
  # Quintile boundaries
  geom_vline(
    xintercept = income_quintiles,
    linetype = "dashed",
    linewidth = 0.8,
    colour = "red"
  ) +
  
  scale_fill_gradientn(
    colours = theme_colors,
    guide = "none"
  ) +
  labs(
    title = "Distribution of Income",
    subtitle = "Dashed lines represent income quintile boundaries",
    x = "Equivalised disposable income",
    y = "Count"
  ) +
  theme_report

basic_histogram
```

The outcome is `1` for a household whose equivalised disposable income
is at or below the 20th percentile, and `0` otherwise.

```{r data-and-outcome}
# Filter
lcfs_data <- lcfs_data[
  is.finite(lcfs_data$eqincdoupdated_eqincdo_updated),
]

income_cutoff <- unname(quantile(
  lcfs_data$eqincdoupdated_eqincdo_updated,
  probs = 0.20 # Change this to adjust the percentile
))

lcfs_data$lowest_income_group <- as.integer(
  lcfs_data$eqincdoupdated_eqincdo_updated <= income_cutoff
)

cat("Lowest-income cutoff:", round(income_cutoff, 2), "\n")
outcome_counts <- table(factor(lcfs_data$lowest_income_group, levels = c(0, 1)))

knitr::kable(data.frame(
  Outcome = c(0, 1),
  Meaning = c("Not in the lowest income quintile", "In the lowest income quintile"),
  Households = as.integer(outcome_counts),
  Percent = round(100 * as.numeric(prop.table(outcome_counts)), 2)
))
```

## Estimate with Selected variables

Income variables are excluded from the predictors because income defines
the outcome.

```{r logistic-regression}
lcfs_data$economic_position_of_hrp <- factor(lcfs_data$economic_position_of_hrp)
lcfs_data$government_office_region <- factor(lcfs_data$government_office_region)
lcfs_data$urban_rural_indicator <- factor(lcfs_data$urban_rural_indicator)

model_variables <- c(
  "lowest_income_group", "household_size", "number_of_children_g019",
  "number_of_workers_in_household_a054", "economic_position_of_hrp",
  "government_office_region", "urban_rural_indicator"
)

# Define the income cutoff before excluding missing predictor values.
model_data <- droplevels(lcfs_data[
  complete.cases(lcfs_data[, model_variables]),
])
cat("Households used in the model:", nrow(model_data), "\n")
cat("Households excluded for missing predictors:",
    nrow(lcfs_data) - nrow(model_data), "\n")

income_model <- glm(
  lowest_income_group ~ household_size + number_of_children_g019 +
    number_of_workers_in_household_a054 + economic_position_of_hrp +
    government_office_region + urban_rural_indicator,
  family = binomial(link = "logit"),
  data = model_data
)
summary(income_model)
```

## Odds ratios

Exponentiation the coefficients gives odds ratios. An odds ratio above 1
indicates higher odds of belonging to the lowest-income group, for
categorical predictors, it is relative to the reference category.

```{r odds-ratios}
cat("Employment reference:", levels(model_data$economic_position_of_hrp)[1], "\n")
cat("Region reference:", levels(model_data$government_office_region)[1], "\n")

odds_ratios <- exp(cbind(
  Odds_ratio = coef(income_model),
  confint.default(income_model)
))
knitr::kable(odds_ratios, digits = 3)
```

## Estimated household probabilities

returns the estimated probability of outcome `1`, between 0 and 1.

```{r predicted-probabilities}
household_predictions <- data.frame(
  lowest_income_group = model_data$lowest_income_group,
  probability_lowest_income = predict(
    income_model, newdata = model_data, type = "response"
  )
)

knitr::kable(head(household_predictions, 10), digits = 3, row.names = FALSE)
```

## Low-income probability by region

Each bar is the average fitted probability for households in that
region, using the full logistic regression model. Households retain
their own characteristics, so the regional differences reflect both
household composition and region; they do not isolate the effect of
region alone.

```{r probability-by-region, fig.width=8, fig.height=6}
region_predictions <- data.frame(
  government_office_region = model_data$government_office_region,
  probability_lowest_income = predict(
    income_model, newdata = model_data, type = "response"
  )
)

region_probabilities <- aggregate(
  probability_lowest_income ~ government_office_region,
  data = region_predictions,
  FUN = mean
)
#Use excel for this
region_probability_plot <- ggplot(
  region_probabilities,
  aes(
    x = probability_lowest_income,
    y = reorder(government_office_region, probability_lowest_income)
  )
) +
  geom_col(fill = "#2E5E7E", width = 0.7) +
  geom_text(
    aes(label = scales::percent(probability_lowest_income, accuracy = 0.1)),
    hjust = -0.15, size = 3.5
  ) +
  scale_x_continuous(
    labels = scales::label_percent(accuracy = 1),
    limits = c(0, NA),
    expand = expansion(mult = c(0, 0.18))
  ) +
  labs(
    title = "Predicted Low-income Probability by Region",
    subtitle = "Average fitted probability across households in each region",
    x = "Probability of belonging to the lowest-income 20%",
    y = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(panel.grid.major.y = element_blank(),
        panel.grid.minor = element_blank())

region_probability_plot
```

## ROC curve and AUC

For performance assessment, fit the same model on 70% of households and
evaluate it on the remaining 30%. Splitting within each outcome group
keeps similar proportions of low-income households in both sets.

```{r roc-test-predictions}
set.seed(123)
roc_train_rows <- unlist(lapply(
  split(seq_len(nrow(model_data)), model_data$lowest_income_group),
  function(rows) sample(rows, size = floor(0.70 * length(rows)))
), use.names = FALSE)

roc_train <- model_data[roc_train_rows, ]
roc_test <- model_data[-roc_train_rows, ]

roc_income_model <- glm(
  formula(income_model),
  family = binomial(link = "logit"),
  data = roc_train
)

roc_test$probability_lowest_income <- predict(
  roc_income_model, newdata = roc_test, type = "response"
)
```

Sensitivity is the proportion of low-income households correctly
identified. Specificity is the proportion of other households correctly
identified. The ROC curve shows their trade-off across probability
thresholds.

```{r roc-curve, fig.width=6, fig.height=6}
lcfs_glm_roc <- pROC::roc(
  response = roc_test$lowest_income_group,
  predictor = roc_test$probability_lowest_income,
  levels = c(0, 1),
  direction = "<", # Higher probabilities indicate the low-income group.
  quiet = TRUE
)

roc_auc <- as.numeric(pROC::auc(lcfs_glm_roc))
roc_auc_ci <- as.numeric(pROC::ci.auc(lcfs_glm_roc, method = "delong"))

knitr::kable(data.frame(
  Training_households = nrow(roc_train),
  Test_households = nrow(roc_test),
  AUC = roc_auc,
  Lower_95_CI = roc_auc_ci[1],
  Upper_95_CI = roc_auc_ci[3]
), digits = 3)

# ggroc's default horizontal axis is specificity, decreasing from 1 to 0.
pROC::ggroc(lcfs_glm_roc) +
  coord_equal() +
  annotate("segment", x = 1, xend = 0, y = 0, yend = 1,
           linetype = "dashed", colour = "grey50") +
  annotate("label", x = 0.90, y = 0.90, label = "Perfect") +
  annotate("label", x = 0.50, y = 0.50, label = "Chance") +
  annotate("label", x = 0.98, y = 0.10,
           label = "Always predict\nnot low-income", hjust = 0, size = 3.5) +
  annotate("label", x = 0.02, y = 0.90,
           label = "Always predict\nlow-income", hjust = 1, size = 3.5) +
  labs(
    title = "LCFS Logistic Regression: Test-set ROC",
    subtitle = paste("AUC =", round(roc_auc, 3)),
    x = "Specificity", y = "Sensitivity"
  ) +
  theme_minimal() #custom theme won't work here
```

## Confusion Matrix

```{r conf-mat}
threshold <- 0.5

actual <- factor(roc_test$lowest_income_group, levels = c(0, 1))

predicted <- factor(
  as.integer(roc_test$probability_lowest_income >= threshold),
  levels = c(0, 1)
)

# Confusion matrix and performance metrics for all test 
classification_results <- caret::confusionMatrix(
  data = predicted,
  reference = actual,
  positive = "1"
)

classification_results
```

# K nearest neighbour

This will predict the **socio-economic group of the household reference
person** using eq household income and food expenditure.

## Prepare and load the data

`socio_economic_group_new`. has been created by grouping 40 original
groups

```{r prep-data}
# Recover the original codes from the existing labelled source variable.
seg_code_lookup <- attr(lcfs_2024$a091, "labels")
names(seg_code_lookup) <- trimws(names(seg_code_lookup))

seg_labels <- trimws(as.character(lcfs_data$socio_economic_group))

lcfs_data <- dplyr::mutate(
    lcfs_data,
    socio_economic_group_num = unname(seg_code_lookup[seg_labels]),

    socio_economic_group_new = dplyr::case_when(
      socio_economic_group_num >= 1  & socio_economic_group_num <= 12 ~ "Managerial & Professional",
      socio_economic_group_num >= 13 & socio_economic_group_num <= 16 ~ "Intermediate Occupations",
      socio_economic_group_num >= 17 & socio_economic_group_num <= 20 ~ "Small Employers & Self-employed",
      socio_economic_group_num >= 21 & socio_economic_group_num <= 23 ~ "Lower Supervisory & Technical",
      socio_economic_group_num >= 24 & socio_economic_group_num <= 35 ~ "Semi-routine & Routine",
      TRUE ~ NA_character_
    ),

    socio_economic_group_new = factor(
      socio_economic_group_new,
      levels = c(
        "Managerial & Professional",
        "Intermediate Occupations",
        "Small Employers & Self-employed",
        "Lower Supervisory & Technical",
        "Semi-routine & Routine"
      )
    ),
    socio_economic_group_num = NULL
  )

table(lcfs_data$socio_economic_group_new, useNA = "ifany")
```

```{r load-data}
lcfs <- data.frame(
  socio_economic_group = lcfs_data$socio_economic_group_new,
  income = lcfs_data$eqincdoupdated_eqincdo_updated,
  food = lcfs_data$food_adults
)
predictors <- c("income", "food")
stopifnot(all(vapply(lcfs[, predictors], is.numeric, logical(1))))

head(lcfs)
summary(lcfs)
```

KNN needs complete numeric predictors and known training labels. exclude
responses outside the five groups.Therefore, Unused factor levels are
dropped before modelling.

```{r complete-observations}
usable <- !is.na(lcfs$socio_economic_group) &
  nzchar(as.character(lcfs$socio_economic_group)) &
  is.finite(lcfs$income) & is.finite(lcfs$food)

cat("Households supplied:", nrow(lcfs), "\n")
cat("Households excluded:", sum(!usable), "\n")
lcfs <- droplevels(lcfs[usable, ])
cat("Households retained:", nrow(lcfs), "\n")
```

## Make training and test sets

Randomly assign about two-thirds of the households **within each group**
to training and the rest to testing. This split keeps every group
represented in both sets.

```{r split-data}
set.seed(123)
if (nlevels(lcfs$socio_economic_group) < 2 ||
    any(table(lcfs$socio_economic_group) < 2)) {
  stop("Need at least two groups and two complete households per group.")
}
training_rows <- unlist(lapply(
  split(seq_len(nrow(lcfs)), lcfs$socio_economic_group),
  function(rows) sample(rows, size = floor(length(rows) * 2 / 3))
), use.names = FALSE)
lcfs_training <- lcfs[training_rows, ]
lcfs_test <- lcfs[-training_rows, ]

c(Training = nrow(lcfs_training), Test = nrow(lcfs_test))
```

```{r group-counts}
group_key <- data.frame(
  Code = sprintf("G%02d", seq_along(levels(lcfs$socio_economic_group))),
  Socio_economic_group = levels(lcfs$socio_economic_group),
  Training = as.integer(table(lcfs_training$socio_economic_group)),
  Test = as.integer(table(lcfs_test$socio_economic_group))
)
knitr::kable(group_key, row.names = FALSE)
```

## Plot the training data

```{r plot-training-data, fig.width=10, fig.height=13}
ggplot(lcfs_training, aes(x = income, y = food)) +
  geom_point(alpha = 0.35, colour = "blue", size = 0.8) +
  facet_wrap(~ socio_economic_group, ncol = 3, labeller = label_wrap_gen(width = 28)) +
  labs(x = "Anonymised household income (GBP)",
       y = "Weekly food spending recorded by adults (GBP)",
       title = "LCFS training households by socio-economic group") +
  theme_report
```

## Scale the predictors

KNN measures how far apart households are. Income has a much wider
numerical range so standardise

```{r scale-predictors}
training_x <- scale(lcfs_training[, predictors])
training_means <- attr(training_x, "scaled:center")
training_sds <- attr(training_x, "scaled:scale")
stopifnot(all(is.finite(training_sds)), all(training_sds > 0))

test_x <- scale(lcfs_test[, predictors],
                center = training_means, scale = training_sds)

knitr::kable(data.frame(
  Predictor = predictors,
  Training_mean = as.numeric(training_means),
  Training_SD = as.numeric(training_sds)
), digits = 2)
```

## Finding nearest-neighbour

```{r inspect-neighbours}
k <- 31
example_row <- which.max(lcfs_test$income)
distances <- sqrt(
  (training_x[, "income"] - test_x[example_row, "income"])^2 +
  (training_x[, "food"] - test_x[example_row, "food"])^2
)

# Include all equally distant households at the kth distance,
# matching class::knn(..., use.all = TRUE).
distance_cutoff <- sort(distances)[k]
neighbour_rows <- which(distances <= distance_cutoff)
neighbour_rows <- neighbour_rows[order(distances[neighbour_rows])]

knitr::kable(lcfs_test[example_row,
                      c("income", "food")],
             row.names = FALSE, caption = "Household to classify")

neighbours <- lcfs_training[neighbour_rows,
                            c("income", "food", "socio_economic_group")]
neighbours$distance <- distances[neighbour_rows]
knitr::kable(neighbours, digits = 3, row.names = FALSE,
             caption = "Nearest training households")
table(droplevels(neighbours$socio_economic_group))
```

Plot to illustrate KNN visualization

```{r plot-neighbours, fig.width=9, fig.height=7}
plot_training <- data.frame(training_x,
                            group_code = factor(group_key$Code[match(
                              lcfs_training$socio_economic_group,
                              group_key$Socio_economic_group
                            )], levels = group_key$Code))
plot_example <- as.data.frame(test_x[example_row, , drop = FALSE])
zoom_radius <- max(0.05, distance_cutoff * 1.5)

ggplot(plot_training, aes(x = income, y = food)) +
  geom_point(aes(colour = group_code), alpha = 0.35) +
  geom_segment(
    data = plot_training[neighbour_rows, ],
    aes(xend = plot_example$income, yend = plot_example$food),
    linewidth = 0.7, colour = "black"
  ) +
  geom_point(data = plot_training[neighbour_rows, ],
             shape = 1, size = 4, colour = "black") +
  geom_point(data = plot_training[neighbour_rows, ],
             aes(colour = group_code), size = 2) +
  geom_point(data = plot_example, shape = 8, size = 4, colour = "black") +
  scale_colour_viridis_d(end = 0.8) +
  coord_equal(
    xlim = plot_example$income + c(-zoom_radius, zoom_radius),
    ylim = plot_example$food + c(-zoom_radius, zoom_radius)
  ) +
  labs(x = "Income (Scaled)", y = "Food expenditure (Scaled)",
       colour = "Socio-economic group", title = "Finding similar households",
       subtitle = "Star: test household; circles and lines: nearest training households") +
  guides(colour = guide_legend(ncol = 6, override.aes = list(alpha = 1))) +
  theme_report +
  theme(legend.position = "bottom")
```

## Predict socio-economic group with KNN

Only `income` and `food` enter the distance calculation.

```{r knn-predictions}
set.seed(123)
knn_predictions <- class::knn(
  train = training_x,
  test = test_x,
  cl = lcfs_training$socio_economic_group,
  k = k,
  use.all = TRUE
)

prediction_examples <- data.frame(
  Actual = lcfs_test$socio_economic_group,
  Predicted = knn_predictions
)
knitr::kable(prediction_examples[example_row, ], row.names = FALSE,
             caption = "Prediction for the illustrated household")
knitr::kable(head(prediction_examples, 10), row.names = FALSE,
             caption = "First ten test predictions")
```

## Confusion matrix

```{r confusion-matrix}
confusion <- table(Actual = lcfs_test$socio_economic_group,
                   Predicted = knn_predictions)
confusion_display <- confusion
dimnames(confusion_display) <- list(Actual = group_key$Code,
                                   Predicted = group_key$Code)
knitr::kable(confusion_display)
```

Accuracy is the proportion of all test households whose group is
classified correctly.

```{r performance}
majority_group <- names(which.max(table(lcfs_training$socio_economic_group)))
accuracy <- mean(knn_predictions == lcfs_test$socio_economic_group)
baseline_accuracy <- mean(lcfs_test$socio_economic_group == majority_group)
group_recall <- diag(confusion) / rowSums(confusion)

cat("Baseline always predicts:", majority_group, "\n")
knitr::kable(data.frame(
  Measure = c("KNN accuracy", "Most-common-group baseline accuracy",
              "Mean recall across groups"),
  Percent = round(100 * c(accuracy, baseline_accuracy,
                         mean(group_recall)), 1)
), row.names = FALSE)

knitr::kable(data.frame(
  Socio_economic_group = rownames(confusion),
  Test_households = as.integer(rowSums(confusion)),
  Correct = as.integer(diag(confusion)),
  Recall_percent = round(100 * group_recall, 1)
), row.names = FALSE)
```

## Hyperparameter Tuning (K)

This gives us K = 17

```{r tune-k}
# 10-fold cross validation, repeated 10 times. 
knn.caret <- train(socio_economic_group~ income + food, 
                   data=lcfs, method="knn", 
                   trControl=trainControl(method="repeatedcv", 
                   number=10, repeats = 5), tuneGrid = expand.grid(k = seq(1,31,2)))

knn.caret

results <- data.frame(knn.caret[4])

head(results)

ggplot(results, aes(x=results.k, y=results.Accuracy)) + 
  geom_line() + labs(x="k", y="Accuracy") +
  theme_report
```

## Predict a group for new households

Replace the example income and food amounts below with new household
values measured in the same way as the training data. Reuse the training
means and standard deviations, the same predictor order, and the same
$k$. These predictions are illustrative and should be read alongside the
held-out performance above.

```{r predict-new-households}
new_households <- data.frame(
  income = c(600, 1200),
  food = c(60, 120)
)
new_x <- scale(new_households[, predictors],
               center = training_means, scale = training_sds)
set.seed(123)
new_households$predicted_socio_economic_group <- class::knn(
  train = training_x,
  test = new_x,
  cl = lcfs_training$socio_economic_group,
  k = k,
  use.all = TRUE
)
knitr::kable(new_households, row.names = FALSE)
```
